---
title: "Introduction to oceanglider"
author: "Dan Kelley (https://orcid.org/0000-0001-7808-5911)"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
    fig_caption: yes
    fig_width: 5
    fig_height: 5
    dpi: 72
    dev.args: list(pointsize=11)
vignette: >
  %\VignetteIndexEntry{Introduction to oceanglider}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

<!-- HOW TO BUILD THE VIGNETTE. -->
<!-- 1. edit this file in vignettes/ -->
<!-- 2. run devtools::build_vignettes() -->
<!-- 3. build package -->


```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

**Abstract.** This vignette explains the basics of using the oceanglider
package on Seaexplorer and Slocum data.

# Introduction

The first step in working with glider data is normally to download them from a
server, and the lack of URL conventions for such servers makes this a
challenging task. The `download.glider()` function is provided to download
glider files from a server, caching the files locally to speed processing.
Using this function demands an understanding of the dataset naming convention
and the server file structure; see the individual examples below, but be aware
that they are only known to work for a few individual datasets.

# Use with seaexplorer glider data

The URL structure in the following example is subject to change without notice.
The portion `SEA024` indicates that this is the glider named 24, and the `M25`
portion refers to the 25-th mission. The pattern, the second argument The
`pld1` in the file pattern requests data from the science stream, the `.2.`
portion indicates that data from the the second descent/ascent sequence (or
'yo') is sought.

```{r}
library(oceanglider)
url <- "ftp://ftp.dfo-mpo.gc.ca/glider/realData/SEA024/M25"
yo2file <- download.glider(url, "pld1.sub.2.gz$", debug=1)
yo2 <- read.glider.seaexplorer(yo2file)
```


# Use with slocum glider data
