---
title: "Introduction to oceanglider"
author: "Dan Kelley (https://orcid.org/0000-0001-7808-5911)"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
    fig_caption: yes
    fig_width: 5
    fig_height: 5
    dpi: 72
    dev.args: list(pointsize=11)
vignette: >
  %\VignetteIndexEntry{Introduction to oceanglider}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- HOW TO BUILD THE VIGNETTE. -->
<!-- 1. edit this file in vignettes/oceanglider.Rmd -->
<!-- 2. devtools::build_vignettes() -->
<!-- 3. devtools::install(build_vignettes=TRUE) -->


```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

**Abstract.** This vignette explains the basics of using the oceanglider
package on Seaexplorer and Slocum data. It is \emph{very much} a work in
progress, because the functions provided by this package are still much under
development.

# Introduction

FIXME: write a couple of paragraphs here about what this package is designed
for, its state of completion, etc.


# Data Sources

The first step in working with glider data is often to download them from a
server. The lack of URL conventions for such servers can make this a somewhat
challenging task to discuss in general terms, and the same can be said about
the variety of file layouts and file names, both across glider types and even
within glider types.

Once data are downloaded, there is an additional task of deciding how to
organize the data locally. At the time of writing, the developers of this
package are leaning towards a scheme with directories named like
```
/data/glider/2019/seaexplorer/sea021/m49/raw
/data/glider/2019/seaexplorer/sea021/m49/sub
```
for seaexplorer data. Here, `sea021` refers to seaexplorer glider number owned
by a particular institute, and `m49` refers to the mission number. The first of
these directories holds "raw" data, i.e. the full datasets as downloaded a
glider after recovery, and the second holds "sub" data, i.e. subsets provided
by the glider during its transmissions while underway on a mission.

A similar scheme might be employed for other glider types, replacing
`seaexplorer` with `slocum`, for example.

Examples of some of these data files are stored within this R package, and are
available using the `system.file` function, as shown throughout this vignette.
This should make it easier for readers to learn how to use this package.


# SeaExplorer Gliders

## Reading files

This package comes with some sample data files that can be used for practice and
illustration.  For example,
```{r}
library(oceanglider)
directory <- system.file("extdata/seaexplorer/sub", package="oceanglider")
```
finds the names of a few selected `"sub"` files (these are files transmitted by
the glider when it is underway; think of the "sub" as an abbreviation for
subset).  Reading these files is done with the
[`read.glider.seaexplorer.realtime`](../help/read.glider.seaexplorer.realtime) function,
e.g.
```{r}
g <- read.glider.seaexplorer.realtime(directory)
```
(when called interactively, this function will print a progress-bar, unless
instructed otherwise by setting the argument \code{progressBar} to
\code{FALSE}.)

The package also comes with a few `"raw"` files, which are larger datasets
downloaded from the glider after recovery.  In the oceanographic convention,
these are called delayed-mode data, so the function to read them is called
[`read.glider.seaexplorer.delayed`](../help/read.glider.seaexplorer.delayed)
```{r}
directory2 <- system.file("extdata/seaexplorer/raw", package="oceanglider")
g2 <- read.glider.seaexplorer.delayed(directory2)
```

Here is how to plot a time-series of pressure from this dataset:
```{r}
plot(g2, which="p")
```

FIXME: why the S4 error here?

(see the help for [`plot,glider-method`](../help/plot-glider-method) for other
plot types).

FIXME: why is only one surfacing? There should be two, I think.

## Summarizing data

The generic `summary()` function has a specialization for glider data, so a
good way to learn about a dataset is with e.g.
```{r}
summary(g)
```

The output of this function is in a form that will be familiar to users of the
`oce` package, with a summary of metadata being given first, then a summary of
the data, and then a listing of some steps in the processing.

Note in particular how variable names have been changed from those in the
original files.

## Examining data

The data are stored in two streams: one, called `glider`, holds data acquired
with sensors on the glider itself, and the other, called `payload1`, holds data
acquired with the first instrumental payload.

The first few lines of the glider stream are given by

```{r}
head(g[["glider"]], 3)
```

The first few lines in the payload are given by
```{r}
head(g[["payload1"]], 3)
```

Actually, in this particular case, the second call could have been
written as
```{r}
head(g[["payload"]], 3)
```
because this glider only had a single payload, However, the first form is
preferable, in terms of code clarity.

It should be noted that the `"salinity"` item is not actually stored in the
data file, but is instead computed (as Practical Salinity) during the reading
process.

## Data extraction

The `[[` notation from the `oce` package is overloaded here. For example,
the pressures in the glider object may be extracted with
`g[["pressure"]]`.

The `NA` values result from the setup of the glider to record only in
particular states of its vertical trajectory. Those states may be found with
`g[["navState"]]`.  (It is not entirely clear why data were recorded at the
first instance of `navState` equaling 100, but not at the following instances.)

**Aside:** the `navState` values are as follows:
105 means the glider is not navigating yet;
115 means the glider is surfacing in preparation for communication;
116 means the glider is at the surface, acquiring a GPS signal, and is communicating;
110 means the glider is inflecting downward;
100 means the glider has ballast set to be descending;
118 means the glider has ballast adjusted to reduce density, so will be inflecting upward;
117 means the glider has ballast set to be ascending.

A common operation is to create a `"ctd"` object, with the `as.ctd` from the
`oce` package, as follows.

FIXME: need to trim in glider-space before making a ctd, I think. And should
split into yos, then plot them in a 3x2 plot I think.

```{r}
library(oce)
ctd <- as.ctd(g[["salinity"]], g[["temperature"]], g[["pressure"]],
              longitude=g[["longitude"]], latitude=g[["latitude"]])
```

The `oce` package has a `summary()` function that provides output that is similar
to that of the present package, e.g.
```{r}
summary(ctd)
```

The `oce` package also provides a way to get a simple TS plot, as follows.

```{r fig.cap="**Figure 1.** CTD plot of a glider yo."}
oce::plotTS(ctd)
```

Note the single very low salinity point.

FIXME: say more about this low S.



In Figure 1, the top left panel shows profiles of salinity and temperature, the
top right panel shows potential density anomaly and the square of buoyancy
frequency, the lower left panel is a temperature-salinity diagram, and the
bottom right panel is a map showing the sampling location (southeast of
Halifax, Nova Scotia).

Note the retracing in some of the panels of Figure 1. To see why, we may use
the `plotScan` function provided by the `oce` package.
```{r fig.cap="**Figure 2.** Pressures in a sample file, with 'Scan' being sequence number of observation."}
plotScan(ctd, type="o")
```

Figure 2 indicates that there are two points near the surface, then a lack of
points during what a descent phase, and then a series of points during the
ascent phase.



# Slocum gliders

**FIXME:** write a few pages here \cdots but this cannot be done until the data
formats are better known, and code is updated.


# References

* Integrated Ocean Observing System (U.S.). “Manual for Quality Control of
Temperature and Salinity Data Observations from Gliders. Version 1.0.” Report.
U.S. Department of Commerce, National Oceanic and Atmospheric Administration,
National Ocean Service, Integrated Ocean Observing System, U.S. Integrated
Ocean Observing System, Glider Data Assembly Center, 2016.
https://www.oceanbestpractices.net/handle/11329/289.

* Integrated Ocean Observing System (U.S.). “NGDAC NetCDF File Format Version
2,” March 27, 2019.
https://github.com/ioos/ioosngdac/wiki/NGDAC-NetCDF-File-Format-Version-2.
