---
title: "Introduction to oceanglider"
author: "Dan Kelley (https://orcid.org/0000-0001-7808-5911)"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
    fig_caption: yes
    fig_width: 5
    fig_height: 5
    dpi: 72
    dev.args: list(pointsize=11)
vignette: >
  %\VignetteIndexEntry{Introduction to oceanglider}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- HOW TO BUILD THE VIGNETTE. -->
<!-- 1. edit this file in vignettes/oceanglider.Rmd -->
<!-- 2. devtools::build_vignettes() -->
<!-- 3. devtools::install(build_vignettes=TRUE) -->


```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

**Abstract.** This vignette explains the basics of using the oceanglider
package on Seaexplorer and Slocum data. It is \emph{very much} a work in
progress, because the functions provided by this package are still much under
development.

# Introduction

The first step in working with glider data is normally to download them from a
server, and the lack of URL conventions for such servers makes this a
challenging task. The `download.glider()` function is provided to download
glider files from a server, caching the files locally to speed processing.
Using this function demands an understanding of the dataset naming convention
and the server file structure; see the individual examples below, but be aware
that they are only known to work for a few individual datasets.

# SeaExplorer Gliders

## Reading files

This package comes with some sample data files that can be used for practice and
illustration.  For example,
```{r}
library(oceanglider)
files <- system.file("extdata/seaexplorer/sub",
                     c("sea021.49.gli.sub.100.gz",
                       "sea021.49.pld1.sub.100.gz"),
                     package="oceanglider")
```
finds the names of two "sub" files (these are files transmitted by the glider
when it is underway; think of the "sub" as an abbreviation for subset).
Reading these files is done with the
[`read.glider.seaexplorer.sub`](../help/read.glider.seaexplorer.sub) function,
e.g.
```{r}
g <- read.glider.seaexplorer.sub(files)
```

Note that the files in the above example have been compressed using gzip, hence
the `".gz"` in the filenames. This is not always the case.  The presence of
`"sub"` in the filenames indicates that these are subset files (i.e. files
transmitted whilst the glider was sampling). Files with `"raw"` in the filename
hold the fuller datasets that are downloaded from the glider after recover, and
such files are read with `read.glider.seaexplorer.raw()`.

## Summarizing data

The generic `summary()` function has a specialization for glider data, so a
good way to learn about a dataset is with e.g.
```{r}
summary(g)
```

The output of this function is in a form that will be familiar to users of the
`oce` package, with a summary of metadata being given first, then a summary of
the data, and then a listing of some steps in the processing.

Note in particular how variable names have been changed from those in the
original files.

## Examining data

The data are stored in two streams: one, called `glider`, holds data acquired
with sensors on the glider itself, and the other, called `payload1`, holds data
acquired with the first instrumental payload.

The first two lines of the glider stream are given by

```{r}
head(g[["glider"]], 3)
```

The first two lines in the payload are given by
```{r}
head(g[["payload1"]], 3)
```

Actually, in this particular case, the second call could have been
written as
```{r}
head(g[["payload"]], 3)
```
because this glider only had a single payload, However, the first form is
preferable, in terms of code clarity.

It should be noted that the `"salinity"` item is not actually stored in the
data file, but is instead computed (as Practical Salinity) during the reading
process.

## Data extraction

The `[[` notation from the `oce` package is overloaded here. For example,
the pressures in the glider object may be extracted with
```{r}
g[["pressure"]]
```
The `NA` values result from the setup of the glider to record only in
particular states of its vertical trajectory. Those states may be found with
```{r}
g[["navState"]]
```
(It is not entirely clear why data were recorded at the first instance of
`navState` equaling 100, but not at the following instances.)

A common operation is to create a `"ctd"` object, with the `as.ctd` from the
`oce` package, as follows.

```{r}
library(oce)
ctd <- as.ctd(g[["salinity"]], g[["temperature"]], g[["pressure"]],
              longitude=g[["longitude"]], latitude=g[["latitude"]])
```

The `oce` package has a `summary()` function that provides output that is similar
to that of the present package, e.g.
```{r}
summary(ctd)
```

The `oce` package also provides a way to get a simple summary plot, as follows.

```{r fig.cap="**Figure 1.** CTD plot of a glider yo."}
oce::plot(ctd)
```

In Figure 1, the top left panel shows profiles of salinity and temperature, the
top right panel shows potential density anomaly and the square of buoyancy
frequency, the lower left panel is a temperature-salinity diagram, and the
bottom right panel is a map showing the sampling location (southeast of
Halifax, Nova Scotia).

Note the retracing in some of the panels of Figure 1. To see why, we may use
the `plotScan` function provided by the `oce` package.
```{r fig.cap="**Figure 2.** Time history of pressure and scan-number for a glider yo. (The scan is a sequence number in the CTD object.)"}
plotScan(ctd, type="o")
```

Figure 2 indicates that there are two points near the surface, then a lack of
points during what a descent phase, and then a series of points during the
ascent phase.

The `oce` package offers a way to isolate the data, using `ctdTrim`; thus, we
can get a standard CTD plot as follows.

```{r fig.cap="**Figure 3.** Trimmed CTD plot, showing just data measured during the upcast."}
oce::plot(ctdTrim(ctd, "upcast"))
```


# Slocum gliders
