---
title: "Introduction to oceanglider"
author: "Dan Kelley (https://orcid.org/0000-0001-7808-5911)"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
    fig_caption: yes
    fig_width: 5
    fig_height: 5
    dpi: 72
    dev.args: list(pointsize=11)
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Introduction to oceanglider}
  %\VignetteEncoding{UTF-8}
---

<!-- HOW TO BUILD THE VIGNETTE. -->
<!-- 1. edit this file in vignettes/ -->
<!-- 2. run devtools::build_vignettes() -->
<!-- 3. build package -->


```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

**Abstract.** This vignette explains the basics of using the oceanglider
package on Seaexplorer and Slocum data.

# Introduction

The first step in working with glider data is normally to download them from a
server, and the lack of URL conventions for such servers makes this a
challenging task. The `download.glider()` function is provided to download
glider files from a server, caching the files locally to speed processing.
Using this function demands an understanding of the dataset naming convention
and the server file structure; see the individual examples below, but be aware
that they are only known to work for a few individual datasets.

# SeaExplorer Gliders

NOTE: The data used in the following are private. The URL structure in the
following example is subject to change without notice.

The portion `SEA024` indicates that this is the glider named 24, and the `M25`
portion refers to the 25-th mission. The pattern, the second argument The
`pld1` in the file pattern requests data from the science stream, the `.2.`
portion indicates that data from the the second descent/ascent sequence (or
'yo') is sought.

```{r}
library(oceanglider)
## url <- "ftp://ftp.dfo-mpo.gc.ca/glider/realData/SEA024/M32"
## filenames <- download.glider(url, "\\.200\\.gz$")
files <- system.file("extdata/seaexplorer/realtime",
                     c("sea024.32.gli.sub.200.gz",
                       "sea024.32.pld1.sub.200.gz"), package="oceanglider")
g <- read.glider.seaexplorer.realtime(files)
```

In applying this example to other situations, it might be helpful to write a
function for generating URLS, e.g. perhaps
```{r}
urlBIO <- function(glider="024", mission=32)
    paste("ftp://ftp.dfo-mpo.gc.ca/glider/realData/", glider, "/M", mission, sep="")
patternBIO <- function(yo=2)
    paste("\\.", yo, "\\.gz$", sep="")
```
where in \code{patternBIO()} using \code{type="pld1"} will yield the 'science'
data, and \code{type="gli"}  will yield the navigation data. (The distinction
between these two types is best explained elsewhere.)

Now, we could get the engineering data for this yo with
```{r eval=FALSE}
filenames <- download.glider(urlBIO("SEA024", 32), patternBIO(200), debug=1)
g <- read.glider.seaexplorer.realtime(filenames)
```
(which is not executed here, owing to server accessibility).

Once we have the data, it can be useful to examine the `glider` and `payload`
components, e.g. with
```{r}
head(g[["glider"]], 3)
head(g[["payload"]], 3)
```

The payload is the preferred data source (consult `help("[[,glider-method")`
for the details of how `[[` works for `glider` objects), so we can make a ctd
object and examine it with
```{r}
ctd <- as.ctd(g[["salinity"]], g[["temperature"]], g[["pressure"]],
              lon=g[["longitude"]], lat=g[["latitude"]])
summary(ctd)
plot(ctd)
```

Note that there seems to be a retracing in the diagrams. To see why, use
```{r}
plotScan(ctd, type="o")
```
which tells us that there is a stray point from a surface interval. We *should* be
able to isolate the upcast by keying on `g[["NAV_RESOURCE"]]==100` but that does not
work, because of this stray point. But, as a stopgap measure, we can use `ctdTrim()` to
get better results:
```{r}
plot(ctdTrim(ctd, "upcast"))
```


# Slocum gliders
