---
title: "Introduction to oceglider"
author: "Dan Kelley (https://orcid.org/0000-0001-7808-5911)"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
    fig_caption: yes
    fig_width: 5
    fig_height: 5
    dpi: 72
    dev.args: list(pointsize=11)
vignette: >
  %\VignetteIndexEntry{Introduction to oceglider}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

**Abstract.** This vignette explains the basics of using the oceglider
package. It is \emph{very much} a work in
progress, because the functions provided by this package are under active
development.

# Introduction

This package is intended to make it easier for analysts to work with data
acquired by ocean gliders. There is a wide variety of glider types and of file
formats, and the purpose of this vignette is to acquaint the reader with a
particular type (the seaexplorer from Alseamar) and particular approach (using
data files produced by the glider itself).

Rather than simply state what might be best practices, I will present a sort of
illustrated diary, showing how I worked my way through a small sample dataset
that is provided with the package.

# Data Sources and file organization

The first step in working with glider data is often to download them from a
server. The lack of URL conventions for such servers can make this a somewhat
challenging task to discuss in general terms, and the same can be said about
the variety of file layouts and file names, both across glider types and even
within glider types.

Once data are downloaded, there is an additional task of deciding how to
organize the data locally. At the time of writing, the developers of this
package are leaning towards a scheme with directories named like
```
/data/glider/2019/seaexplorer/sea021/m49/raw
/data/glider/2019/seaexplorer/sea021/m49/sub
```
for seaexplorer data. Here, `sea021` refers to seaexplorer glider number owned
by a particular institute, and `m49` refers to the mission number. The first of
these directories holds "raw" data, i.e. the full datasets as downloaded a
glider after recovery, and the second holds "sub" data, i.e. subsets provided
by the glider during its transmissions while underway on a mission.


Examples of some of these data files are stored within this R package, and are
available using the `system.file` function, as shown throughout this vignette.
This should make it easier for readers to learn how to use this package.


# SeaExplorer Gliders

This package comes with some sample data files that can be used for practice and
illustration.  For example,
```{r}
library(oceglider)
directory <- system.file("extdata/seaexplorer/sub", package="oceglider")
```
finds the names of a few selected `"sub"` files (these are files transmitted by
the glider when it is underway; think of the "sub" as an abbreviation for
subset).  Reading these files is done with
[`read.glider.seaexplorer.realtime`](../help/read.glider.seaexplorer.realtime) function,
e.g.
```{r}
g <- read.glider.seaexplorer.realtime(directory)
```
(when called interactively, this function will print a progress-bar, unless
instructed otherwise by setting the argument \code{progressBar} to
\code{FALSE}.)

The package also comes with a few `"raw"` files, which are larger datasets
downloaded from the glider after recovery.  In the oceanographic convention,
these are called delayed-mode data, so the function to read them is called
[`read.glider.seaexplorer.delayed`](../help/read.glider.seaexplorer.delayed)
```{r}
directory2 <- system.file("extdata/seaexplorer/raw", package="oceglider")
g2 <- read.glider.seaexplorer.delayed(directory2, level=1)
```

Here is how to plot a time-series of pressure from this dataset:
```{r}
plot(g2, which="p")
```

Here is how to see locations in space:
```{r}
plot(g2, which="map")
```

The points are on a line because
[`read.glider.seaexplorer.delayed`](../help/read.glider.seaexplorer.delayed)
interpolates longitude and latitude between surfacings, and this small subset
of glider data was selected to have a position record at the first and last of
the sequence.

It is also noteworthy that some of the spots seem to have only a few data. This
can be checked with
```{r}
table(g2[["yoNumber"]])
```

the result of which show that some `yoNumber` values contain too few data to be
considered profiles. There are several ways to explore this further, but for
the present purpose, let's make a new object that contains only those yos that
have more than, say, 5 points:
```{r}
g3 <- subset(g2, yolength >= 5)
```
As a check on that idea, note that
```{r}
unique(g3[["yoNumber"]])
```
verifies that we got the desired yos.

Quite often, it will be desirable to work with individual yos. With only 3, as in this
small test case, we might as well plot them (say as temperature profiles). For that, it
is helpful to split by yo number, and then cycle though the subsets, converting to CTD
objects and plotting with functions from the \CRANpkg{oce} package.

The individual yos can be examined by splitting the `payload` portion:
```{r}
payload <- g3[["payload"]] # or [["payload1"]]
s <- split(payload, payload$yoNumber)
par(mfrow=c(2, 3)) # we happen to know there are 3 significant yos
for (ss in s) {
    ctd <- oce::as.ctd(salinity=ss[["salinity"]],
        temperature=ss[["temperature"]],
        pressure=ss[["pressure"]],
        longitude=ss[["longitude"]],
        latitude=ss[["latitude"]])
   oce::plotProfile(ctd, xtype="salinity+temperature", eos="unesco")
}
```

(Here, the UNESCO equation of state is used, because the newer TEOS10/GSW version requires
longitude and latitude information to be valid, which is not the case in the
present version of the package at the end of a dataset, owing to how
[`read.glider.seaexplorer.delayed`](../help/read.glider.seaexplorer.delayed) is
works.)

Note how there is a line connecting near-surface and deep data. To learn more
about that, let's use an \CRANpkg{oce} function to create a ``scan'' plot, say
with the last of these `ctd` objects just created.
```{r}
plotScan(ctd, type="p")
```

This makes it clear that the yo contains near-surface data, then skips to
depth, and then returns to the surface. A more careful examination with
```{r}
par(mfrow=c(1, 2))
plot(s[[1]][["time"]], s[[1]][["pressure"]], xlab="Time", ylab="Pressure")
plot(s[[1]][["time"]], s[[1]][["navState"]], xlab="Time", ylab="navState")
```

reveals that if the desire was to obtain a simple upcast, we might consider
subsetting by the value of the `navState` variable.

Noting that the `navState` values are as follows:
* 105 means the glider is not navigating yet;
* 115 means the glider is surfacing in preparation for communication;
* 116 means the glider is at the surface, acquiring a GPS signal, and is communicating;
* 110 means the glider is inflecting downward;
* 100 means the glider has ballast set to be descending;
* 118 means the glider has ballast adjusted to reduce density, so will be inflecting upward;
* 117 means the glider has ballast set to be ascending.
it may be sensible to select for 117, if the desire is to focus on the ascent
portion of the yo. So, retracing our steps, we have


```{r}
g4 <- subset(g3, navState==117)
payload <- g4[["payload"]] # or [["payload1"]]
s <- split(payload, payload$yoNumber)
par(mfrow=c(2, 2)) # we happen to know there are 3 significant yos
for (ss in s) {
    ctd <- oce::as.ctd(salinity=ss[["salinity"]],
        temperature=ss[["temperature"]],
        pressure=ss[["pressure"]],
        longitude=ss[["longitude"]],
        latitude=ss[["latitude"]])
    oce::plotProfile(ctd, xtype="salinity+temperature", eos="unesco")
}
```

These results are not much better! Selecting by `navState` is clearly not
sufficient. However, `oce` provides the `ctdTrim()` function for
isolating the ascent (called `"upcast"` in the CTD context), e.g.
```{r}
par(mfrow=c(2, 2)) # we happen to know there are 3 significant yos
for (ss in s) {
    ctd <- oce::as.ctd(salinity=ss[["salinity"]],
        temperature=ss[["temperature"]],
        pressure=ss[["pressure"]],
        longitude=ss[["longitude"]],
        latitude=ss[["latitude"]])
   ctd <- oce::ctdTrim(ctd, "upcast")
   oce::plotProfile(ctd, xtype="salinity+temperature", eos="unesco")
}
```

This looks reasonable now. Let's try a waterfall temperature plot, with black
for the first yo, red for the second, green for the third, and blue for the
fourth, as explained in the legend.
```{r}
par(mfrow=c(1, 1))
for (i in seq_along(s)) {
    ss <- s[[i]]
    ctd <- oce::as.ctd(salinity=ss[["salinity"]],
        temperature=ss[["temperature"]],
        pressure=ss[["pressure"]],
        longitude=ss[["longitude"]],
        latitude=ss[["latitude"]])
    ctd <- oce::ctdTrim(ctd, "upcast")
    if (i == 1) {
       oce::plotProfile(ctd, xtype="temperature", eos="unesco")
    } else {
       lines(ctd[["temperature"]], ctd[["pressure"]], col=i)
    }
}
legend("bottomright", lwd=1, col=seq_along(s),
    legend=paste("yo",
        as.character(lapply(s,function(ss) ss$yoNumber[1]))))
```

At this stage, readers will likely have some ideas for doing more with this
sample dataset, and ought to be aware that many of the steps here might not
be appropriate for other datasets, e.g. those in which data were acquired on
both ascent and descent, or only on descent.


# References

* Integrated Ocean Observing System (U.S.). “Manual for Quality Control of
Temperature and Salinity Data Observations from Gliders. Version 1.0.” Report.
U.S. Department of Commerce, National Oceanic and Atmospheric Administration,
National Ocean Service, Integrated Ocean Observing System, U.S. Integrated
Ocean Observing System, Glider Data Assembly Center, 2016.
https://www.oceanbestpractices.net/handle/11329/289.

* Integrated Ocean Observing System (U.S.). “NGDAC NetCDF File Format Version
2,” March 27, 2019.
https://github.com/ioos/ioosngdac/wiki/NGDAC-NetCDF-File-Format-Version-2.
