---
title: "Handling Slocum Glider Data"
author: "Dan Kelley (https://orcid.org/0000-0001-7808-5911)"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
    fig_caption: yes
    fig_width: 5
    fig_height: 5
    dpi: 72
    dev.args: list(pointsize=11)
vignette: >
  %\VignetteIndexEntry{Handling Slocum Glider Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

**Abstract.** This vignette explains the basics of using the oceglider package
on Slocum data. The main difference from SeaExplorer files is in the layout of
the objects, with SeaExplorer basically considering glider and payload in
separate streams. But normal use, with plotting, subsetting, etc., does not
require the user to consider the details of the layout.

# Create Temporary Directory for CRAN Submission

For this vignette to be suitable for CRAN submission, it must not download
files to the user's working space.  We avoid that by setting aside space in the
next line.

```{r}
fileNC <- tempfile(fileext = ".nc") # to hold a binary (NetCDF) file
```

Note that this `fileNC` file will be deleted at the end of this vignette.
Obviously, in actual practice, the analyst is likely to store the downloaded
file in a more permanent location, and not to delete it when finished a
particular task.

# Some Common Steps with Slocum Datasets

We will use a dataset from CPROOF (Reference 1).

```{r}
library(oceglider)
url <- paste0(
    "https://cproof.uvic.ca/gliderdata/deployments/",
    "dfo-bumblebee998/dfo-bumblebee998-20221207/",
    "L0-timeseries/",
    "dfo-bumblebee998-20221207_delayed.nc"
)
file <- gsub(".*/", "", url)
# 45.8 MB file: download takes 0.4s user, 0.9s system but 11s elapsed
if (!file.exists(file)) {
    system.time(curl::curl_download(url = url, destfile = file, mode = "wb"))
}
g <- read.glider.netcdf(file) # under 0.2s
```

It always makes sense to get a summary, to see what data are contained
in the object.

```{r}
summary(g)
```

Next, make a map to see where the sampling was done.

```{r}
plot(g, which = "map")
```

This map comports with the one on Reference 2, except we have proper scaling of
the axes, because `oceglider` computes an aspect ratio that will preserve
shapes at the centre of the plot.

Next, examine the pressure-time trajectory.

```{r}
plot(g, which = "p", cex = 0.2)
```

This looks good.  Let's colour-code by temperature.

```{r}
plot(g, which = "p", type = "p", pch = 20, colorby = "temperature")
```

Notice that there is very little red in the diagram.  Close inspection will
reveal that the high temperatures corresponding to this colour are at the start
of the time series.  The methods illustrated in the SeaExplorer vignette can be
employed here, but for brevity we will just use percentile trimming to get a
diagram that at least provides an initial view of the data.

```{r}
plot(g,
    which = "p", type = "p", pch = 20, colorby = "temperature",
    colorbylim = quantile(g[["temperature"]], c(0.01, 0.99), na.rm = TRUE)
)
```

Readers might agree that this version looks better, displaying
more features than in the first plot.

Salinity is often a problem, but we'll do the same thing, first plotting all
the data.

```{r}
plot(g, which = "p", type = "p", pch = 20, colorby = "salinity")
```

And, second, trimming.

```{r}
plot(g,
    which = "p", type = "p", pch = 20, colorby = "salinity",
    colorbylim = quantile(g[["salinity"]], c(0.01, 0.99), na.rm = TRUE)
)
```

Again, we see improvement.

At this stage, we have some confidence that data are being read properly and
that plotting, extraction with `[[` and some other basic features work.
Readers are directed to the SeaExplorer vignette for more ideas on things that
might be worth pursuing.

## Cleanup of Temporary Files

As noted at the outset, CRAN rules dictate that the downloaded material be
deleted.  In actual analysis, though, it is unlikely that such material would
be deleted, so don't just copy this line!

```{r}
unlink(fileNC) # where the binary (NetCDF) file went
```

# References

1. “dfo-bumblebee998-2022-12-07 Time series Delayed” Accessed September 5,
   2024.
<https://cproof.uvic.ca/gliderdata/deployments/dfo-bumblebee998/dfo-bumblebee998-20221207/>


