---
title: "Handling Slocum Glider Data"
author: "Dan Kelley (https://orcid.org/0000-0001-7808-5911)"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
    fig_caption: yes
    fig_width: 5
    fig_height: 5
    dpi: 72
    dev.args: list(pointsize=11)
vignette: >
  %\VignetteIndexEntry{Handling Slocum Glider Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

**Abstract.** This vignette explains the basics of using the oceglider package
on Slocum data. The main difference from SeaExplorer files is in the layout of
the objects, with SeaExplorer basically considering glider and payload in
separate streams. But normal use, with plotting, subsetting, etc., does not
require the user to consider the details of the layout.

For this vignette to be suitable for CRAN submission, it must not download
files to the user's working space.  We avoid that by setting aside space in the
next line.

```{r}
fileNC <- tempfile(fileext = ".nc") # to hold a binary (NetCDF) file
```

Note that this `fileNC` file will be deleted at the end of this vignette.
Obviously, in actual practice, the analyst is likely to store the downloaded
file in a more permanent location, and not to delete it when finished a
particular task.

# Some Common Steps with Slocum Datasets

We will use a dataset from CPROOF (Reference 1).

```{r}
library(oceglider)
url <- paste0(
    "https://cproof.uvic.ca/gliderdata/deployments/",
    "dfo-bumblebee998/dfo-bumblebee998-20221207/",
    "L0-timeseries/",
    "dfo-bumblebee998-20221207_delayed.nc"
)
file <- gsub(".*/", "", url)
# 45.8 MB file: download takes 0.4s user, 0.9s system but 11s elapsed
if (!file.exists(file)) {
    system.time(curl::curl_download(url = url, destfile = file, mode = "wb"))
}
g <- read.glider.netcdf(file) # under 0.2s
```

It always makes sense to get a summary, to see what data are contained
in the object.

```{r}
summary(g)
```

Next, make a map to see where the sampling was done.

```{r}
plot(g, which = "map")
```

This map comports with the one on Reference 2, except we have proper scaling of
the axes, because `oceglider` computes an aspect ratio that will preserve
shapes at the centre of the plot.

Next, examine the pressure-time trajectory.

```{r}
plot(g, which = "p")
```

Note that there is a large region of whitespace on the right of the plot panel.
This is because the dataset has time values in that interval, but all the
pressure values are NA. The `subset()` function could be used to trim
the dataset, but that is not necessarily wise because there could be other
valuable data (even just longitude and latitude are valuable, in case the
hydrographic sensors provide nothing useable).

We can show temperature on a similar plot, by switching from lines to points,
and colour-coding by temperature.

```{r}
plot(g, which = "p", type = "p", pch = 20, cex = 0.2, colorby = "temperature")
```

Notice that the colour palette ranges up to red, although only a small region
of the graph has that colour. but a quick glance at the graph does not reveal
many points that are of sufficient temperature to be indicated with that color.
The methods illustrated in the SeaExplorer vignette can be employed here, but
for brevity we will just use percentile trimming to get a diagram that at least
provides an initial view of the data.

```{r}
plot(g,
    which = "p", type = "p", pch = 20, cex = 0.2, colorby = "temperature",
    colorbylim = quantile(g[["temperature"]], c(0.01, 0.99), na.rm = TRUE)
)
```

Readers might agree that this version looks better, displaying
more features than in the first plot.

Salinity is often a problem, but we'll do the same thing, first plotting all
the data.

```{r}
plot(g, which = "p", type = "p", pch = 20, cex = 0.2, colorby = "salinity")
```

And, second, trimming.

```{r}
plot(g,
    which = "p", type = "p", pch = 20, cex = 0.2, colorby = "salinity",
    colorbylim = quantile(g[["salinity"]], c(0.01, 0.99), na.rm = TRUE)
)
```

Again, we see improvement.

At this stage, we have some confidence that data are being read properly and
that plotting, extraction with `[[` and some other basic features work.
Readers are directed to the SeaExplorer vignette for more ideas on things that
might be worth pursuing.

As noted at the outset, CRAN rules dictate that the downloaded material be
deleted.  In actual analysis, though, it is unlikely that such material would
be deleted, so don't just copy this line!

```{r}
unlink(fileNC) # where the binary (NetCDF) file went
```

# References

1. “dfo-bumblebee998-2022-12-07 Time series Delayed” Accessed September 5,
   2024.
<https://cproof.uvic.ca/gliderdata/deployments/dfo-bumblebee998/dfo-bumblebee998-20221207/>


